# GSEA

### Setup environment

```{r}
library(DESeq2)
library(magrittr)
```

### Get ginger vs *A. thaliana* blastp results

Ginger protein -\> *A. thaliana* protein

```{r}
blastp_res <- read.table(
  file = "outputs/blastp/ginger_vs_ath.tsv",
  header = FALSE,
  sep = "\t"
)
head(blastp_res)

```

### Get protein ID -\> gene ID relationship for ginger and *A. thaliana*

#### Define GTF utils

Define parser for GTF attributes

```{r}
parse_attribute <- function(string) {
  tmp <- string %>%
    trimws(., which = "both") %>% # 去除字串開頭與/或結尾的空白字元
    strsplit(., ";") %>%
    unlist %>% # 攤平成向量
    trimws(., which = "both") %>% #再去掉頭尾空白
    strsplit(., " ") # 以空格切分成 key/value 向量
  out <- c()
  for (i in tmp) {
    out[i[1]] = paste(i[-1], collapse = " ") # i[-1]：除了第一個以外的所有元素組成一個字串向量，也就是value各段。
  }
  return(out)
}
```

GTF header

```{r}
gtf_header <- c( # 定義讀入 GTF 檔案時要用的欄位名稱
  "seqname",
  "source",
  "feature",
  "start",
  "end",
  "score",
  "strand",
  "frame",
  "attributes"
)
```

1.  Ginger

```{r}
# import gtf
ginger_gtf <- read.table(
  file = "references/ncbi_dataset/data/GCF_018446385.1/genomic.finished.gtf",
  sep = "\t",
  header = FALSE,
  col.names = gtf_header,
  comment.char = "#"
)

# find every first CDS lines of each protein coding gene
cds_lines <- ginger_gtf$feature == "CDS"
every_first_cds <- 
  (cds_lines & c(FALSE, !cds_lines[-length(cds_lines)])) %>%  #判斷前一列是不是CDS
  which #回傳所有邏輯向量

# get protein ID and gene ID from parsed attributes
ginger_p2g <- matrix(
  ncol = 2,
  nrow = length(every_first_cds)
)
j <- 1
for (nth_row in every_first_cds) {
  attrs <- parse_attribute(ginger_gtf$attributes[nth_row])
  ginger_p2g[j, 1] <- attrs["protein_id"]
  ginger_p2g[j, 2] <- attrs["gene_id"]
  j <- j + 1
}

# set row name of the matrix as protein ID
rownames(ginger_p2g) <- ginger_p2g[, 1]
colnames(ginger_p2g) <- c("pid", "gid")
```

2.  *A. thaliana*

```{r}
ath_annot_info <- read.table(
  file = "references/ath/Phytozome/PhytozomeV9/Athaliana/annotation/Athaliana_167_annotation_info.txt",
  header = TRUE,
  sep = "\t",
  comment.char = ""
)
head(ath_annot_info)
ath_p2g <- ath_annot_info[, c("peptideName", "locusName")] %>% as.matrix
colnames(ath_p2g) <- c("pid", "gid")
rownames(ath_p2g) <- ath_p2g[, 1]
```

### Get Batch 2 DESeq2 results

```{r}
dds <- readRDS(file = "outputs/rmd-output/deseq.Batch2.RDS")
resnames <- DESeq2::resultsNames(dds)
print(resnames)
```







#### Group "condition_LETC_vs_CK"

Get DEGs

```{r}
FDR <- 0.05
FCThreashold <- 1.5
name <- "condition_LETC_vs_CK"
res <- results(
    dds,
    alpha = FDR,
    lfcThreshold = FCThreashold,
    name = name
)

# print results
res
```

#找DEG上下調基因
```{r} 
DEGs_up <- subset(
  res,
  (res$padj < 0.05) & (res$log2FoldChange > 1.5)
) %>% rownames
DEGs_dn <- subset(
  res,
  (res$padj < 0.05) & (res$log2FoldChange < -1.5)
) %>% rownames
```

Export filtered DEGs: \|log2 fold change\| \> 1.5; adjusted p-value \< 0.05

```{r}
ggids <- sapply(blastp_res$V1, function(x){ # blastp_res 只有蛋白質對蛋白質的對應，不包含基因資訊，GSEA 要的是基因 ID不是蛋白 ID，所以要對回去原本的結果找出其他資訊
  if (x %in% rownames(ginger_p2g)) {
    return(ginger_p2g[x, 2])
  } else {
    return("Unmatched")
  }
})
blastp_res$ginger_gid <- ggids
blastp_res$ath_gid <- sapply(
  blastp_res$V2, function(x) {
    strsplit(x, split = "[.]") %>% unlist %>% extract(1)
  }) %>% 
  unname
blastp_res <- blastp_res[order(blastp_res$V11), ]
```

```{r}
where_l2fc <- abs(res$log2FoldChange) > 1.5
where_adjp <- res$padj < 0.05

# filter NA #如果 log2FoldChange 或 padj 是 NA，就當作不顯著處理，防止錯誤
where_l2fc[is.na(where_l2fc)] <- FALSE
where_adjp[is.na(where_adjp)] <- FALSE

# filtered
filtered_res <- res[where_adjp & where_l2fc, ]

# sorted
final_res <- filtered_res[order(filtered_res$padj), ]

# get corresponding Ath gene ID
agids <- rownames(final_res) %>% 
  sapply(., function(x){ #在 %>% 中的 . 是「目前傳入的物件」
    blastp_res$ath_gid[which(blastp_res$ginger_gid == x)[1]] #which(...)[1]：只取第一個 match的index，避免重複對應。
  })
final_res$ath_best_hit <- agids

# export result
write.csv(
  final_res[final_res$log2FoldChange > 0,],
  file = sprintf("outputs/rmd-output/DEG-table/DEG.up.%s.csv", name),
  row.names = TRUE,
  quote = FALSE
)
write.csv(
  final_res[final_res$log2FoldChange < 0,],
  file = sprintf("outputs/rmd-output/DEG-table/DEG.dn.%s.csv", name),
  row.names = TRUE,
  quote = FALSE
)
```

Match Ginger DEG gene ID to *A. thaliana* gene ID

1.  Up-regulated

```{r}
curated_gp2g <- subset(ginger_p2g, (ginger_p2g[, 2] %in% DEGs_up))
curated_blastp_res <- subset(blastp_res, blastp_res$V1 %in% curated_gp2g[, 1])
curated_blastp_res$apid <- curated_blastp_res$V2 %>% sapply(
  function(x){strsplit(x, split = "[|]") %>% unlist %>% extract(1)}
)
# curated_ap2g <- subset(ath_p2g, ath_p2g[, 1] %in% curated_blastp_res$apid)
# ggids <- curated_blastp_res$V1 %>% 
#   sapply(.,
#     function(x) {
#       curated_gp2g[x, 2]
#     }
#   )
agids_up <- curated_blastp_res$apid %>%
  sapply(.,
    function(x) {
      strsplit(x, split = "[.]") %>% unlist %>% extract(1)
    }
  ) %>% 
  unname
cat(
  agids_up,
  sep = "\n",
  file = sprintf("outputs/rmd-output/%s.up.txt", name)
)
```

Results retrieved from PlantGSEA

```{bash}
wget \
  --directory-prefix=outputs/PlantGSEA/condition_RKN_vs_Water/up/ \
  http://structuralbiology.cau.edu.cn/PlantGSEA/tmp/362703666.sorted_detail2
```

2.  Down-regulated

```{r}
curated_gp2g <- subset(ginger_p2g, (ginger_p2g[, 2] %in% DEGs_dn))
curated_blastp_res <- subset(blastp_res, blastp_res$V1 %in% curated_gp2g[, 1])
curated_blastp_res$apid <- curated_blastp_res$V2 %>% sapply(
  function(x){strsplit(x, split = "[|]") %>% unlist %>% extract(1)}
)
# curated_ap2g <- subset(ath_p2g, ath_p2g[, 1] %in% curated_blastp_res$apid)
# ggids <- curated_blastp_res$V1 %>% 
#   sapply(.,
#     function(x) {
#       curated_gp2g[x, 2]
#     }
#   )
agids_dn <- curated_blastp_res$apid %>%
  sapply(.,
    function(x) {
      strsplit(x, split = "[.]") %>% unlist %>% extract(1)
    }
  ) %>% 
  unname
cat(
  agids_dn,
  sep = "\n",
  file = sprintf("outputs/rmd-output/%s.dn.txt", name)
)
```

```{bash}
wget \
  --directory-prefix=outputs/PlantGSEA/condition_RKN_vs_Water/down/ \
  http://structuralbiology.cau.edu.cn/PlantGSEA/tmp/252656794.sorted_detail2
```

#### Group "condition_LITR_vs_CK"

Get DEGs

```{r}
FDR <- 0.05
FCThreashold <- 1.5
name <- "condition_LITR_vs_CK"
res <- results(
    dds,
    alpha = FDR,
    lfcThreshold = FCThreashold,
    name = name
)

# print results
res
```

#找DEG上下調基因
```{r} 
DEGs_up <- subset(
  res,
  (res$padj < 0.05) & (res$log2FoldChange > 1.5)
) %>% rownames
DEGs_dn <- subset(
  res,
  (res$padj < 0.05) & (res$log2FoldChange < -1.5)
) %>% rownames
```

Export filtered DEGs: \|log2 fold change\| \> 1.5; adjusted p-value \< 0.05

```{r}
ggids <- sapply(blastp_res$V1, function(x){ # blastp_res 只有蛋白質對蛋白質的對應，不包含基因資訊，GSEA 要的是基因 ID不是蛋白 ID，所以要對回去原本的結果找出其他資訊
  if (x %in% rownames(ginger_p2g)) {
    return(ginger_p2g[x, 2])
  } else {
    return("Unmatched")
  }
})
blastp_res$ginger_gid <- ggids
blastp_res$ath_gid <- sapply(
  blastp_res$V2, function(x) {
    strsplit(x, split = "[.]") %>% unlist %>% extract(1)
  }) %>% 
  unname
blastp_res <- blastp_res[order(blastp_res$V11), ]
```

```{r}
where_l2fc <- abs(res$log2FoldChange) > 1.5
where_adjp <- res$padj < 0.05

# filter NA #如果 log2FoldChange 或 padj 是 NA，就當作不顯著處理，防止錯誤
where_l2fc[is.na(where_l2fc)] <- FALSE
where_adjp[is.na(where_adjp)] <- FALSE

# filtered
filtered_res <- res[where_adjp & where_l2fc, ]

# sorted
final_res <- filtered_res[order(filtered_res$padj), ]

# get corresponding Ath gene ID
agids <- rownames(final_res) %>% 
  sapply(., function(x){ #在 %>% 中的 . 是「目前傳入的物件」
    blastp_res$ath_gid[which(blastp_res$ginger_gid == x)[1]] #which(...)[1]：只取第一個 match的index，避免重複對應。
  })
final_res$ath_best_hit <- agids

# export result
write.csv(
  final_res[final_res$log2FoldChange > 0,],
  file = sprintf("outputs/rmd-output/DEG-table/DEG.up.%s.csv", name),
  row.names = TRUE,
  quote = FALSE
)
write.csv(
  final_res[final_res$log2FoldChange < 0,],
  file = sprintf("outputs/rmd-output/DEG-table/DEG.dn.%s.csv", name),
  row.names = TRUE,
  quote = FALSE
)
```

Match Ginger DEG gene ID to *A. thaliana* gene ID

1.  Up-regulated

```{r}
curated_gp2g <- subset(ginger_p2g, (ginger_p2g[, 2] %in% DEGs_up))
curated_blastp_res <- subset(blastp_res, blastp_res$V1 %in% curated_gp2g[, 1])
curated_blastp_res$apid <- curated_blastp_res$V2 %>% sapply(
  function(x){strsplit(x, split = "[|]") %>% unlist %>% extract(1)}
)
# curated_ap2g <- subset(ath_p2g, ath_p2g[, 1] %in% curated_blastp_res$apid)
# ggids <- curated_blastp_res$V1 %>% 
#   sapply(.,
#     function(x) {
#       curated_gp2g[x, 2]
#     }
#   )
agids_up <- curated_blastp_res$apid %>%
  sapply(.,
    function(x) {
      strsplit(x, split = "[.]") %>% unlist %>% extract(1)
    }
  ) %>% 
  unname
cat(
  agids_up,
  sep = "\n",
  file = sprintf("outputs/rmd-output/%s.up.txt", name)
)
```

Results retrieved from PlantGSEA

```{bash}
上面file為空檔案
```

2.  Down-regulated

```{r}
curated_gp2g <- subset(ginger_p2g, (ginger_p2g[, 2] %in% DEGs_dn))
curated_blastp_res <- subset(blastp_res, blastp_res$V1 %in% curated_gp2g[, 1])
curated_blastp_res$apid <- curated_blastp_res$V2 %>% sapply(
  function(x){strsplit(x, split = "[|]") %>% unlist %>% extract(1)}
)
# curated_ap2g <- subset(ath_p2g, ath_p2g[, 1] %in% curated_blastp_res$apid)
# ggids <- curated_blastp_res$V1 %>% 
#   sapply(.,
#     function(x) {
#       curated_gp2g[x, 2]
#     }
#   )
agids_dn <- curated_blastp_res$apid %>%
  sapply(.,
    function(x) {
      strsplit(x, split = "[.]") %>% unlist %>% extract(1)
    }
  ) %>% 
  unname
cat(
  agids_dn,
  sep = "\n",
  file = sprintf("outputs/rmd-output/%s.dn.txt", name)
)
```

```{bash}
wget \
  --directory-prefix=outputs/PlantGSEA/condition_RKN_vs_Water/down/ \
  http://structuralbiology.cau.edu.cn/PlantGSEA/tmp/224849810.sorted_detail2
```